Stateless Model Checking with Data-Race Preemption Points

### 2015-11-14 pldi-16/evaluation.tex
clean up stale TODOs

% TODO: Abbreviate SMC, and fixed-PP approach.

@@

	\item Does \quicksand~improve upon state-of-the-art MC?
		\begin{enumerate}
			\item Does Iterative Deepening find bugs faster
				%than SSS-MC
				in subset state spaces, even without data-race PPs?
			% Probably not... ICB is state of the art here.
			% In large tests, can Iterative Deepening provide partial verification by completing smaller state spaces
			\item Do data-race PPs expose new bugs that couldn't be found with SSS-MC's fixed-PP-set approach?
				% Elaborate later:
				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
		\end{enumerate}
	\item Does MC improve the accuracy of data-race detection?
		\begin{enumerate}
			\item Do we avoid false positives compared to a single-execution data race analysis?
				% Explain later as:
				% How many data-race candidates were verified as benign
				% But to be fair, you have to count how many DRs are reported as "couldn't test these, check yourself" at the end.
				% Also Include:
				% How many false positives does the free-re-malloc technique suppress?
				% TODO: If you have time, re-run all of the dr-only bug tests, with DR_FALSE_NEG enabled, and see how much fewer bugs get found (how many bugs get pushed past the time limit?)
			% TODO: This one's optional. You can give up on it.
			\item Do we find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
				%TODO: for this experiment, set EXPLORE_BACKWARDS=0
				% TODO: And disable false-neg malloc-free technique
				% TODO: And also disable the confirmed/suspected thing
				% (where mem.c waits for reorder observed before
				% messaging the latter half of the DR to QS)
		\end{enumerate}
\end{enumerate}

@@

\subsection{Test Suite}
% TODO: Maybe say how many lines of code total? How many lines on average per P2/pintos? (careful with pintos; lots of basecode)
Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
%across the Spring and Fall 2014 and Spring 2015 semesters;
and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
%
The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and r/w locks;
all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
%
The Pintos kernel project
involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
using provided bare-bones mutex, context-switch, and virtal memory implementations
\cite{pintos}.
% P2 SLOC stats: 1807 avg; 1723 median; range 1181-4114.
% All numbers, obtained with:
% cd p2s; for i in */*; do wc -l $i/user/libthread/*.{c,h,S} $i/user/libthread/*/*.{c,h,S} ; done | grep total
% 1181 1192 1221 1230 1238 1240 1243 1261 1275 1307 1310 1318 1325 1334 1336 1345 1366
% 1388 1388 1403 1415 1416 1430 1451 1478 1498 1527 1589 1618 1635 1638 1654 1675 1676
% 1716 1719 1720 1723 1723 1727 1737 1743 1744 1751 1769 1777 1782 1789 1789 1812 1918
% 1926 1946 1994 2022 2043 2066 2077 2088 2099 2131 2164 2172 2190 2215 2227 2277 2282
% 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
Though not ``real world'' programs, both projects are quite complex:
the P2s average 1807 lines of code,%C and x86 assembly (stddev 489.5),
% Pintos SLOC stats: 718.1923077 avg; 643 median; range 68-1821
% Obtained with:
%function getloc {
%	diff -ru /tmp/shit/pintoses/di/src/$2/ $1/src/$2 | diffstat | grep insertions | sed 's/.*changed, //' | sed 's/ insertion.*//'
%}
%for i in `ls -d /tmp/shit/pintoses/* | grep -v -- "-p1$"`; do
%#for i in /tmp/shit/pintoses/daniel-deniz*; do
%	echo -ne "$i";
%	tloc=`getloc $i threads`
%	uloc=`getloc $i userprog`
%	dloc=`getloc $i devices`
%	echo -e "\t$tloc\t$uloc\t$dloc"
%done
and the Pintoses average 718.

@@

\newcommand\mxtest{\texttt{mx\_test}}
\newcommand\tej{\texttt{join\_test}}
\newcommand\bct{\texttt{bcast\_test}}
\newcommand\paraguay{\texttt{signal\_test}}
\newcommand\paradise{\texttt{sem\_test}}
\newcommand\rwldgr{\texttt{rwlock\_test}}
We tested P2s with 6 multithreaded programs:
% from the 410 test suite % XXX: I would like to say this but this is a lie; figure out what else i can say instead
% each tailored to exercise a different part of the P2 project
\mxtest, for locking algorithm correctness,
\tej, a test of thread lifecycle,
\bct~and \paraguay~for cvars,
\paradise~for semaphores,
and \rwldgr~for r/w locks.
For \mxtest, \paradise, and \paraguay, we used {\tt without\_function} to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
and for \mxtest~we enabled \landslide's mutex-testing option
(see \sect{\ref{sec:landslide}}).
\newcommand\prisema{\texttt{sched\_test}}
\newcommand\waitsimple{\texttt{wait\_test}}
\newcommand\alarmsimul{\texttt{alarm\_test}}
% TODO: Add more test cases
We tested the Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
\alarmsimul, for the timer sleep routine,
and \waitsimple, for process lifecycle system calls\footnote{
	Some of the Pintoses were partially implemented, so each test could only be run on a subset of the 78 submissions; see Table~\ref{tab:drbugs}.
}
The source code of all 9 tests is available at
{\em [submitted as supplementary material]}.
For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} lock.
Finally, all tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
In total, the test suite comprises 629 unique state spaces,
at least 165 of which contain bugs.

@@

% TODO: If you have time, rerun all the quicksand experiments JUST running the 4 base state spaces. Give it 10/4 hours of cpu budget on 4 cores. John's expt.

@@

% TODO: Make a table with a breakdown by test case.

@@

%Figure~\ref{fig:bargraph}
The left half of Table~\ref{tab:drbugs}
breaks down the number and types of bugs found by each test case.
% The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
%TODO: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
% (as well as on the mutex API boundaries).
% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
even compared to the other test cases\footnote{
% "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
	The one bug SSS-MC found was in a fully-assembly lock implementation. {\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions. Preempting only on {\tt yield()} (in the contention loop) was sufficient.}.
%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
%Though for many applications of MC, assuming a correct lock implementation is sufficient,
Though it suffices for many applications of MC to assume correctly-implemented locks,
we consider this strong evidence that any verification of low-level synchronization code must incorporate data-race PPs.

@@

Among the 69 true data-races which initially fit the malloc-recycle pattern,
none of them exposed a new bug when used as a PP.
This shows that for other data-race tools,
suppressing malloc-recycle candidates may be a productive heuristic,
even if unsound without Iterative Deepening.
However, \quicksand~was able to correctly identify 69 violations of the heuristic among 30 tests,
and fall back to classifying them with DPOR,
which we consider a much stronger verification.

@@

% TODO: Make a CDF graph of how many tests got total verification here.

@@

% TODO: You should introduce this in the narrative as a key contribution way earlier.
\subsection{Finding nondeterministic data-race candidates}
\label{sec:eval-falseneg}
Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.
In such cases, a single-pass dynamic data-race detector
%could not achieve the coverage necessary
would fail
to identify a racing access pair as a candidate to begin with.
%
We counted how many such data-races, used as PPs, led to \quicksand~finding new bugs,
thereby making them {\em false negatives} of the single-pass approach.
We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).

@@

At first, it might seem unfair to compare the data race bugs \quicksand~finds (10 CPU-hours)
against the candidates found by a prior work's single execution (less than a minute).
We argue this comparison is meaningful because prior work data-race tools %, being not integrated with a MC,
are not intended to discover new data race candidates under subsequent runs. %before discovering any initial candidates.
Running a data-race tool repeatedly for 10 CPU-hours could potentially uncover some of these false negatives at random,
but stress testing's comparative problem with achieving reliable coverage is already well-understood
% TODO: 1 more good citation here
\cite{chess-icb,gambit}.
This experiment also suggests that
other replay-based data-race tools \cite{racefuzzer,portend} should be configured to use the same sort of feedback loop as Iterative Deepening (\sect{\ref{sec:related-dr}}).
%i.e., discovering transitively-reachable data races while testing initial ones.
%although that would still not simultaneously be able to provide total verifications.
% TODO: read gambit paper

>>>>

@@

	\item Does \quicksand~improve upon state-of-the-art MC?
		\begin{enumerate}
			\item Does Iterative Deepening find bugs faster
				%than SSS-MC
				in subset state spaces, even without data-race PPs?
			% Probably not... ICB is state of the art here.
			% In large tests, can Iterative Deepening provide partial verification by completing smaller state spaces
			\item Do data-race PPs expose new bugs that couldn't be found with SSS-MC's fixed-PP-set approach?
				% Elaborate later:
				% Among those, how many were missed in a {\em completed} execution of the otherwise ``maximal'' state space?
		\end{enumerate}
	\item Does MC improve the accuracy of data-race detection?
		\begin{enumerate}
			\item Do we avoid false positives compared to a single-execution data race analysis?
				% Explain later as:
				% How many data-race candidates were verified as benign
				% But to be fair, you have to count how many DRs are reported as "couldn't test these, check yourself" at the end.
				% Also Include:
				% How many false positives does the free-re-malloc technique suppress?
				% to do (done) If you have time, re-run all of the dr-only bug tests, with DR_FALSE_NEG enabled, and see how much fewer bugs get found (how many bugs get pushed past the time limit?)
				% Answer: Just 1. (for p2s at least; as of time of writing, pintos dr-falsenegs not run yet)
			\item Do we find data-race bugs that would be false-negatives during a single-execution analysis?%Do we avoid false negatives compared to single-pass?
		\end{enumerate}
\end{enumerate}

@@

\subsection{Test Suite}
Our test suite consists of \numthrlibs~``P2'' student thread libraries, from CMU's 15-410 OS class,
and \numpintoses~``Pintos'' student kernels, from Berkeley's CS162 and U. Chicago's CS230 OS classes.
%
The P2 thread library comprises \texttt{thr\_create()}, \texttt{thr\_exit()}, \texttt{thr\_join()}, mutexes, condition variables, semaphores, and r/w locks;
all implemented from scratch in userspace with a UNIX-like system call interface \cite{kspec,thrlib}.
%
The Pintos kernel project
involves implementing priority scheduling, \texttt{sleep()}, and user-space process management (\texttt{wait()} and \texttt{exit()})
using provided bare-bones mutex, context-switch, and virtal memory implementations
\cite{pintos}.
% P2 SLOC stats: 1807 avg; 1723 median; range 1181-4114.
% All numbers, obtained with:
% cd p2s; for i in */*; do wc -l $i/user/libthread/*.{c,h,S} $i/user/libthread/*/*.{c,h,S} ; done | grep total
% 1181 1192 1221 1230 1238 1240 1243 1261 1275 1307 1310 1318 1325 1334 1336 1345 1366
% 1388 1388 1403 1415 1416 1430 1451 1478 1498 1527 1589 1618 1635 1638 1654 1675 1676
% 1716 1719 1720 1723 1723 1727 1737 1743 1744 1751 1769 1777 1782 1789 1789 1812 1918
% 1926 1946 1994 2022 2043 2066 2077 2088 2099 2131 2164 2172 2190 2215 2227 2277 2282
% 2384 2387 2483 2486 2503 2514 2551 2597 2610 2665 4114
Though not ``real world'' programs, both projects are quite complex:
the P2s average 1807 lines of code,%C and x86 assembly (stddev 489.5),
% Pintos SLOC stats: 718.1923077 avg; 643 median; range 68-1821
% Obtained with:
%function getloc {
%	diff -ru /tmp/shit/pintoses/di/src/$2/ $1/src/$2 | diffstat | grep insertions | sed 's/.*changed, //' | sed 's/ insertion.*//'
%}
%for i in `ls -d /tmp/shit/pintoses/* | grep -v -- "-p1$"`; do
%#for i in /tmp/shit/pintoses/daniel-deniz*; do
%	echo -ne "$i";
%	tloc=`getloc $i threads`
%	uloc=`getloc $i userprog`
%	dloc=`getloc $i devices`
%	echo -e "\t$tloc\t$uloc\t$dloc"
%done
and the Pintoses average 718.

@@

\newcommand\mxtest{\texttt{mx\_test}}
\newcommand\tej{\texttt{join\_test}}
\newcommand\bct{\texttt{bcast\_test}}
\newcommand\paraguay{\texttt{signal\_test}}
\newcommand\paradise{\texttt{sem\_test}}
\newcommand\rwldgr{\texttt{rwlock\_test}}
We tested P2s with 6 multithreaded programs:
% from the 410 test suite % XXX: I would like to say this but this is a lie; figure out what else i can say instead
% each tailored to exercise a different part of the P2 project
\mxtest, for locking algorithm correctness,
\tej, a test of thread lifecycle,
\bct~and \paraguay~for cvars,
\paradise~for semaphores,
and \rwldgr~for r/w locks.
For \mxtest, \paradise, and \paraguay, we used {\tt without\_function} to blacklist {\tt thr\_create}, {\tt thr\_exit}, and {\tt thr\_join},
and for \mxtest~we enabled \landslide's mutex-testing option
(see \sect{\ref{sec:landslide}}).
\newcommand\prisema{\texttt{sched\_test}}
\newcommand\waitsimple{\texttt{wait\_test}}
\newcommand\alarmsimul{\texttt{alarm\_test}}
% TODO CAMREADY: Add another test case
We tested the Pintoses with 3 programs from the class test suite: \prisema, a test of the kernel scheduling algorithm,
\alarmsimul, for the timer sleep routine,
and \waitsimple, for process lifecycle system calls\footnote{
	Some of the Pintoses were partially implemented, so each test could only be run on a subset of the 78 submissions; see Table~\ref{tab:drbugs}.
}
The source code of all 9 tests is available at
{\em [submitted as supplementary material]}.
For all tests, we used {\tt without\_function} to blacklist PPs on the {\tt malloc} lock.
Finally, all tests were run on 12-core 3.2 GHz Xeon machines with 12GB of RAM.
In total, the test suite comprises 629 unique state spaces,
at least 165 of which contain bugs.

@@

@@

@@

%Figure~\ref{fig:bargraph}
The left half of Table~\ref{tab:drbugs}
breaks down the number and types of bugs found by each test case.
% The astute observer will note that the sum of all types of failures observed is more than the total number of data-race bugs...
%TODO CAMREADY: Run a mutex expt where "all atomic instrs" are PPs. See how many bugs are missed anyway.
% Wwe re-ran the \mxtest control experiment with \landslide~hard-coded to preempt on any atomic instruction
% (as well as on the mutex API boundaries).
% Still, this smarter configuration for SSS-MC found only 99999999999 bugs of \quicksand's 13.
In \mxtest~in particular, the control experiment found dramatically fewer bugs (just 1),
even compared to the other test cases\footnote{
% "Aren't all lock impls assembly?" "Yes, but this one was ALL assembly."
	The one bug SSS-MC found was in a fully-assembly lock implementation. {\tt yield()}'s return value clobbered a value stored in {\tt \%eax}, which could lead to a failure after two repeated contentions. Preempting only on {\tt yield()} (in the contention loop) was sufficient.}.
%Intuitively, this is due to our control experiment being able to preempt only on the boundaries of the API which
%Though for many applications of MC, assuming a correct lock implementation is sufficient,
Though it suffices for many applications of MC to assume correctly-implemented locks,
we consider this strong evidence that any verification of low-level synchronization code must incorporate data-race PPs.

@@

Among the 69 true data-races which initially fit the malloc-recycle pattern,
none of them exposed a new bug when used as a PP.
This shows that for other data-race tools,
suppressing malloc-recycle candidates may be a productive heuristic,
even if unsound without Iterative Deepening.
However, \quicksand~was able to correctly identify 69 violations of the heuristic among 30 distinct tests,
and fall back to classifying them with DPOR,
which we consider a much stronger verification.

@@

@@

\subsection{Finding nondeterministic data-race candidates}
\label{sec:eval-falseneg}
Some memory accesses may be hidden in a control flow path that requires a nondeterministic preemption to be executed.
In such cases, a single-pass dynamic data-race detector
%could not achieve the coverage necessary
would fail
to identify a racing access pair as a candidate to begin with.
%
We counted how many such data-races, used as PPs, led to \quicksand~finding new bugs,
thereby making them {\em false negatives} of the single-pass approach.
We classified each data-race candidate according to whether \landslide~had reported them during the first interleaving, before any backtracking or preempting: if so, they were {\em deterministic data races} (hence could be found by single-pass).

@@

At first, it might seem unfair to compare the data race bugs \quicksand~finds (10 CPU-hours)
against the candidates found by a prior work's single execution (less than a minute).
We argue this comparison is meaningful because prior work data-race tools %, being not integrated with a MC,
are not intended to discover new data race candidates under subsequent runs. %before discovering any initial candidates.
Running a data-race tool repeatedly for 10 CPU-hours could potentially uncover some of these false negatives at random,
but stress testing's comparative problem with achieving reliable coverage is already well-understood
\cite{chess-icb,gambit}.
This experiment also suggests that
other replay-based data-race tools \cite{racefuzzer,portend} should be configured to use the same sort of feedback loop as Iterative Deepening (\sect{\ref{sec:related-dr}}).
%i.e., discovering transitively-reachable data races while testing initial ones.
%although that would still not simultaneously be able to provide total verifications.
% TODO: read gambit paper

### 2015-11-14 pldi-16/paper.tex
implement joshua's revisions pt1
touch up di a bit
clean up stale TODOs

% TODO: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
\newcommand\landslide{\textsc{Landslide}}
\newcommand\quicksand{\textsc{Quicksand}}
\newcommand\simics{\textsc{Simics}}
\newcommand{\sect}[1]{\S #1}
\newcommand\hilight[2]{\color{#1}#2\color{black}}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{brickred}{RGB}{192,0,0}

@@

We present an {\em Iterative Deepening} framework for stateless model checking,
which manages the exploration of many state spaces using different subsets of preemption points.
It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
and it incorporates a data-race analysis to dynamically add new preemption points to classify each data race as buggy or benign.
%
Our evaluation shows that combining data-race detection with stateless model checking can find new bugs neither technique would find on their own.
%Our evaluation shows this technique is
%more effective than single-state-space model checking
%%both at finding more bugs and at completing more state spaces when no bug exists.
%both at finding the same bugs faster and at finding new bugs entirely.

@@

Our evaluation shows that \quicksand~achieves better bug-finding results than either single-pass data-race detection or single-state-space model checking (SSS-MC) alone.
%We dramatically reduce false-positive data race reports,
%%by verifying the absence of bugs in their associated state spaces
%and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
% TODO: Not true; Replace instead with a claim about partial verification.
Compared to SSS-MC, we find bugs faster in smaller subset state spaces,
%when the maximal state space is too large to explore
and find new bugs with data-race PPs that the maximal state space would not expose at all.
Moreover, for tests where all state spaces with data-race PPs can be completed, we provide an equally strong verification as if every single instruction had been used as a PP.

>>>>

% TODO: Don't forget to run e.g. s/\\quicksand /\\quicksand~/
\newcommand\landslide{\textsc{Landslide}}
\newcommand\quicksand{\textsc{Quicksand}}
\newcommand\simics{\textsc{Simics}}
\newcommand{\sect}[1]{\S #1}
\newcommand\hilight[2]{\color{#1}#2\color{black}}
\definecolor{olivegreen}{RGB}{0,127,0}
\definecolor{brickred}{RGB}{192,0,0}
\definecolor{commentblue}{RGB}{0,0,255}

@@

We present an {\em Iterative Deepening} framework for stateless model checking,
which manages the exploration of many state spaces using different subsets of preemption points.
It uses state space estimation to prioritize jobs most likely to complete in a fixed CPU budget,
and it incorporates a data-race analysis to dynamically add new preemption points to classify each data race as buggy or benign.
%
Our evaluation shows that combining data-race detection with stateless model checking can find new bugs that neither technique would find on their own.
%Our evaluation shows this technique is
%more effective than single-state-space model checking
%%both at finding more bugs and at completing more state spaces when no bug exists.
%both at finding the same bugs faster and at finding new bugs entirely.

@@

Our evaluation shows that \quicksand~achieves better bug-finding results than either single-pass data-race detection or single-state-space model checking (SSS-MC) alone.
%We dramatically reduce false-positive data race reports,
%%by verifying the absence of bugs in their associated state spaces
%and find bugs caused by nondeterministic data-races in alternate thread interleavings missed entirely by single-pass analysis.
Compared to SSS-MC,
%, we find bugs faster in smaller subset state spaces, %% not true, or at least, not by very much
%when the maximal state space is too large to explore
we find new bugs with data-race PPs that the maximal state space would not expose at all.
Moreover, for tests where all state spaces with data-race PPs can be completed, we provide an equally strong verification as if every single instruction had been used as a PP.
